---
title: "DSC-607 Data Mining: Clustering Models"
author: "Michael Albers"
date: '2022-06-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

- Discuss the theoretical background of cluster analyses (in general) in your program comments.

- Perform cluster analysis of your data using two different methods and interpret and discuss the results.

- How do the results of the two methods differ? Why?

- What statistic did you use to compare the performance of your two clustering algorithms?



## Data Set

This data set contains information on New York City air quality surveillance data. The agency is Department of Health and Mental Hygiene (DOHMH). The data set was downloaded from https://data.cityofnewyork.us/Environment/Air-Quality/c3uy-2p5r.

Air pollution is one of the most important environmental threats to urban populations and while all people are exposed, pollutant emissions, levels of exposure, and population vulnerability vary across neighborhoods. Exposures to common air pollutants have been linked to respiratory and cardiovascular diseases, cancers, and premature deaths. These indicators provide a perspective across time and NYC geographies to better characterize air quality and health in NYC.




## Clustering Background



```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(skimr)
library(janitor)
```

```{r}
aq <- read.csv('data/Air_Quality.csv')
```

```{r}
glimpse(aq)
```

```{r}
aq_clean <- aq %>% clean_names()
```

```{r}
colnames(aq_clean)
```

```{r}

```

